---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# metaprep

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Meta-analyzing effect estimates from multi-arm trials requires accounting for the dependence that arises when several treatment arms are compared to the same control group. The methods are well understood and `metafor::rma.mv()` implements them: you pass a block-diagonal variance-covariance matrix that encodes the dependence. The annoying part is building that matrix and keeping it synchronized with the estimates as you filter, group, and subset.

`metaprep` handles the bookkeeping. It provides an `estimates_vcov` object that keeps coefficient estimates and their variance-covariance matrix in sync through dplyr operations, plus thin wrappers around `metafor` that read from it directly.

## Installation

Install the development version from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("acoppock/metaprep")
```

## Example

Prepare a fit from each trial, bind them together, and build a synchronized `estimates_vcov` object:

```{r example, message=FALSE}
library(dplyr)
library(metaprep)

set.seed(343)
fit_1 <- lm(Y ~ Z, data = data.frame(Y = rnorm(50),  Z = sample(c("T0", "T1"), 50, TRUE)))
fit_2 <- lm(Y ~ Z, data = data.frame(Y = rnorm(100), Z = sample(c("T0", "T1", "T2"), 100, TRUE)))

fits_df <- bind_rows(
  `Study 1` = prep_fit(fit_1, term = "ZT1"),
  `Study 2` = prep_fit(fit_2, term = c("ZT1", "ZT2")),
  .id = "study"
)

ev <- as_estimates_vcov(fits_df)
ev
```

dplyr verbs keep the vcov matrix in sync with the estimates:

```{r}
ev |> filter(study == "Study 2")
```

Meta-analyze with the `metafor` wrapper, which pulls the estimates and vcov from the object automatically:

```{r meta, eval=requireNamespace("metafor", quietly = TRUE)}
library(metafor)
ev |> rma_mv_helper(yi = estimate, random = ~ 1 | id)
```

See `vignette("multi-arm")` for the full workflow, including component extraction, grouped meta-analyses, and what happens under the hood.
