---
title: "Preparing multi-arm trials for meta-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing multi-arm trials for meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When meta-analyzing effect estimates obtained from multi-arm trials, we have to account for the dependence across estimates that arises when multiple treatment arms are compared to the same control group. The statistical methods for this have been worked out, as has an excellent software implementation in the `metafor` package. In short, we can pass a block-diagonal variance-covariance matrix to `rma.mv()` that appropriately accounts for this dependence.

The annoying part is making the block-diagonal variance-covariance matrix and keeping it synchronized with the estimates data frame. The `metaprep` package helps with this bookkeeping by providing an `estimates_vcov` object that keeps estimates and their variance-covariance matrix in sync through dplyr operations.

```{r setup, message=FALSE}
set.seed(343)
library(dplyr)
library(randomizr)
library(estimatr)
library(metaprep)
```

# Three experiments

Let's make fake data for three experiments. All three have a continuous outcome. The first is small (N = 50) with two arms. The second is larger (N = 100) with three arms. The third is the biggest (N = 200) with four arms. We use `randomizr::complete_ra()` to assign the treatment arm `Z`, which performs complete random assignment and automatically labels the arms `T1`, `T2`, and so on.

```{r}
dat_1 <- data.frame(Z = complete_ra(50, num_arms = 2), Y = rnorm(50))
dat_2 <- data.frame(Z = complete_ra(100, num_arms = 3), Y = rnorm(100))
dat_3 <- data.frame(Z = complete_ra(200, num_arms = 4), Y = rnorm(200))
```

Now we estimate treatment effects with `estimatr::lm_robust()`, which returns heteroskedasticity-robust standard errors:

```{r}
fit_1 <- lm_robust(Y ~ Z, data = dat_1)
fit_2 <- lm_robust(Y ~ Z, data = dat_2)
fit_3 <- lm_robust(Y ~ Z, data = dat_3)
```

# Basic workflow: estimates_vcov objects

Here comes the `metaprep` part. We prep each fit, bind them together, and create an `estimates_vcov` object:

```{r}
prepped_fit_1 <- prep_fit(fit_1, term = "ZT2")
prepped_fit_2 <- prep_fit(fit_2, term = c("ZT2", "ZT3"))
prepped_fit_3 <- prep_fit(fit_3, term = c("ZT2", "ZT3", "ZT4"))

fits_df <- bind_rows(
  `Study 1` = prepped_fit_1,
  `Study 2` = prepped_fit_2,
  `Study 3` = prepped_fit_3,
  .id = "study"
)

ev <- as_estimates_vcov(fits_df)
ev
```

The `estimates_vcov` object keeps the estimates and variance-covariance matrix synchronized. You can use dplyr verbs on it:

```{r}
ev |> filter(study == "Study 2")
ev |> mutate(abs_estimate = abs(estimate))
ev |> arrange(desc(estimate))
```

Now we're ready to do meta-analysis using `rma_mv_helper()`:

```{r eval=requireNamespace("metafor", quietly = TRUE)}
library(metafor)
library(broom)
meta_fit <- ev |> rma_mv_helper(yi = estimate, random = ~ 1 | id)
tidy(meta_fit)
```

The `rma_mv_helper()` function is a convenience wrapper that automatically passes the estimates data frame and vcov matrix to `metafor::rma.mv()`. The `random = ~ 1 | id` argument specifies a random effect for each estimate (identified by the `id` column that `estimates_vcov` objects automatically include).

# Cluster-robust standard errors

Meta-analyses of multi-arm trials usually want standard errors clustered by study, since arms from the same study share variation the block-diagonal vcov does not capture. Pass a `cluster` column and `rma_mv_helper()` wraps the fit in `metafor::robust()` for you, so you get cluster-robust standard errors in one step:

```{r eval=requireNamespace("metafor", quietly = TRUE) && requireNamespace("clubSandwich", quietly = TRUE)}
ev |>
  rma_mv_helper(yi = estimate, random = ~ 1 | id, cluster = study) |>
  tidy(conf.int = TRUE)
```

`cluster = study` is evaluated in the estimates, exactly like `yi`. By default `clubSandwich = TRUE` requests CR2 standard errors; set it to `FALSE` for metafor's CR0 estimator.

# Meta-regression with moderators

To ask whether effects vary with a study-level or arm-level characteristic, add the moderator to the object with `mutate()` and pass it through `mods`. Keeping the moderator on the `estimates_vcov` object, rather than on a detached data frame, is what keeps it aligned with the vcov:

```{r eval=requireNamespace("metafor", quietly = TRUE)}
ev |>
  mutate(big_study = study == "Study 3") |>
  rma_mv_helper(yi = estimate, mods = ~ big_study, random = ~ 1 | id) |>
  tidy(conf.int = TRUE)
```

If a `mods` formula names a column that is not on the object, `rma_mv_helper()` stops with an error instead of silently dropping the moderator, so a variable that never made it onto the object cannot quietly produce an empty regression.

## Reading the pooled fit

`rma_mv_helper()` returns a standard `metafor` object (or a `robust.rma` object when `cluster` is supplied), so `broom::tidy()` and `broom::glance()` read it directly, with no need to pull slots out by hand:

```{r eval=requireNamespace("metafor", quietly = TRUE)}
glance(meta_fit)
```

`glance()` gives the scalar model summary and, for `rma.uni` or structured `rma.mv` fits, the heterogeneity `tau^2`. For a multilevel `rma.mv` (for example `random = list(~ 1 | study, ~ 1 | id)`) the variance components live in `meta_fit$sigma2`, which `glance()` does not expand. That matches broom's convention for mixed models, where per-group variances are not folded into the one-row `glance()` summary.

# Alternative workflow: extract components separately

If you prefer to work with the components separately, you can extract them:

```{r}
estimates_df <- get_estimates_df(fits_df)
estimates_df

vcov_matrix <- get_vcov(fits_df)
vcov_matrix
```

These functions work on both the prepped fits data frame and on `estimates_vcov` objects:

```{r}
estimates_df <- get_estimates_df(ev)
vcov_matrix <- get_vcov(ev)
```

You can also extract model-level summaries with `get_glance_df()`:

```{r}
glance_df <- get_glance_df(fits_df)
glance_df
```

Then conduct meta-analysis the traditional way:

```{r eval=requireNamespace("metafor", quietly = TRUE)}
meta_fit_traditional <- rma.mv(
  yi = estimate, V = vcov_matrix, data = estimates_df, random = ~ 1 | id
)
tidy(meta_fit_traditional)
```

# Grouping and nesting workflows

The `estimates_vcov` object supports `nest_by()` for grouped meta-analyses:

```{r eval=requireNamespace("metafor", quietly = TRUE)}
ev_grouped <- ev |>
  mutate(arm_type = if_else(term %in% c("ZT2"), "Treatment 2", "Treatment 3or4"))

grouped_results <- ev_grouped |>
  nest_by(arm_type) |>
  mutate(
    meta_fit = list(rma_mv_helper(data, yi = estimate, random = ~ 1 | id)),
    meta_results = list(tidy(meta_fit))
  ) |>
  tidyr::unnest(meta_results)

grouped_results |> select(arm_type, estimate, std.error, p.value)
```

# Combining and transforming objects

If you built several `estimates_vcov` objects separately, for example one per data source, `bind_estimates_vcov()` stacks them into one, assembling a block-diagonal vcov with zero covariance between objects:

```{r}
ev_a <- as_estimates_vcov(bind_rows(`Study 1` = prepped_fit_1, .id = "study"))
ev_b <- as_estimates_vcov(bind_rows(`Study 2` = prepped_fit_2, `Study 3` = prepped_fit_3, .id = "study"))
bind_estimates_vcov(ev_a, ev_b)
```

To change the sign or the units of the estimates, use `rescale_estimates_vcov()`, not `mutate()`. The dplyr methods keep the vcov row-aligned but do not transform it, so `mutate(estimate = -estimate)` would flip the estimates while leaving the covariances inconsistent. `rescale_estimates_vcov()` updates the vcov to match (\(V \mapsto \mathrm{diag}(s)\,V\,\mathrm{diag}(s)\)), which is what you want for a sign flip (`by = +1`/`-1`, including a partial flip of some arms only) or a unit change:

```{r}
ev |> rescale_estimates_vcov(by = 100)
```

# Correlated estimates with no single regression: bootstrapping

Multi-arm trials are the easy case, because one regression produces all the arms and `vcov()` hands us their covariances. Sometimes estimates are correlated because they share subjects, but no single regression produces them both. A common design: every subject takes a survey experiment, and a random third of them also take a lab experiment. The survey effect and the lab effect are estimated on overlapping samples, so they covary, but they have different outcomes and different assignment variables and will not stack into one model.

The fix is to bootstrap the whole design -- resampling *subjects*, so that the shared-sample dependence is what the replicates reproduce -- and read the covariance off the replicate estimates with `cov()`. `make_estimates_vcov()` then wraps those hand-built pieces into an ordinary `estimates_vcov` object.

```{r}
n <- 400
overlap_dat <- data.frame(
  Z_survey = rbinom(n, 1, 0.5),
  in_lab = rbinom(n, 1, 1 / 3)
)
overlap_dat$Z_lab <- if_else(overlap_dat$in_lab == 1, rbinom(n, 1, 0.5), NA)
overlap_dat$Y_survey <- 0.2 * overlap_dat$Z_survey + rnorm(n)
overlap_dat$Y_lab <- 0.5 * overlap_dat$Z_lab + 0.6 * overlap_dat$Y_survey + rnorm(n)

estimate_both <- function(d) {
  c(
    survey = coef(lm_robust(Y_survey ~ Z_survey, data = d))[["Z_survey"]],
    lab = coef(lm_robust(Y_lab ~ Z_lab, data = filter(d, in_lab == 1)))[["Z_lab"]]
  )
}
```

Each iteration draws a bootstrap sample of subjects and re-estimates both effects on it. Resampling row indices of `overlap_dat` resamples *subjects*: a subject drawn into the replicate brings both their survey outcome and, if they were in the lab third, their lab outcome, which is what carries the shared-sample dependence into the replicate estimates.

```{r}
sims <- 200
boots <- matrix(NA_real_, nrow = sims, ncol = 2,
                dimnames = list(NULL, c("survey", "lab")))

for (i in 1:sims) {
  boot_indices <- sample(n, n, replace = TRUE)
  boot_dat <- overlap_dat[boot_indices, ]
  boots[i, ] <- estimate_both(boot_dat)
}

head(boots)
```

The covariance of the replicates is the covariance of the estimates. The point estimates come from the real data, not from the bootstrap:

```{r}
V_boot <- cov(boots)
point <- estimate_both(overlap_dat)
```

The estimates data frame is built by hand, in the same row order as `V_boot`. Include `std.error` so that `rescale_estimates_vcov()` has standard errors to rescale:

```{r}
boot_estimates <- data.frame(
  study = "Overlapping samples",
  term = names(point),
  estimate = point,
  std.error = sqrt(diag(V_boot))
)

ev_boot <- make_estimates_vcov(boot_estimates, V_boot)
ev_boot
get_vcov(ev_boot)
```

The off-diagonal element is the shared-subject covariance, which no single regression would have given us.

The result is an ordinary `estimates_vcov` object, so it binds with objects built the usual way and pools the same way. The bootstrapped covariance is carried into the combined matrix, with zeros against the studies it shares no subjects with:

```{r}
combined <- bind_estimates_vcov(ev_boot, ev)
round(get_vcov(combined), 4)
```

```{r eval=requireNamespace("metafor", quietly = TRUE)}
combined |> rma_mv_helper(yi = estimate, random = ~ 1 | study)
```

Two things to keep in mind. The vcov is matched to the estimates **by position**, so row `i` of the estimates data frame must be row and column `i` of the matrix; any dimnames on the matrix are discarded and replaced by the object's `id`. Building both from the same ordered vector, as `estimate_both()` does above, makes that automatic. And `make_estimates_vcov()` checks that the matrix is square and symmetric, erroring on asymmetry beyond floating-point noise -- a genuinely asymmetric matrix means the rows and columns are misaligned, which would make the meta-analysis silently wrong.

# What's going on under the hood

`prep_fit()` conducts three post-estimation operations on the regression fit: `tidy()`, `glance()`, and `vcov()`. It subsets the rows of the tidy object and the rows and columns of the vcov object to the terms relevant for the meta-analysis -- in this case excluding the intercept.

The resulting three objects are wrapped in a tibble where the cells are themselves data frames:

```{r}
prepped_fit_1
```

We then bind all the prepped fits together with an id variable:

```{r}
fits_df
```

The `get_estimates_df()` function extracts and unnests the estimates:

```{r}
get_estimates_df(fits_df)
```

And the `get_vcov()` function extracts each `vcov_obj` and combines them into a block-diagonal matrix:

```{r}
get_vcov(fits_df)
```

The `as_estimates_vcov()` function creates a special object that keeps these two components synchronized:

```{r}
ev <- as_estimates_vcov(fits_df)
class(ev)
```

This object has an `id` column that links each row of estimates to the corresponding row/column of the vcov matrix:

```{r}
ev$estimates$id
rownames(ev$vcov)
```

When you use dplyr verbs on an `estimates_vcov` object, both the estimates and vcov are updated together:

```{r}
ev_study2 <- ev |> filter(study == "Study 2")
nrow(get_estimates_df(ev_study2))
dim(get_vcov(ev_study2))
```

This synchronization is what makes the `estimates_vcov` object useful for meta-analysis workflows involving filtering, grouping, and subsetting.
