---
title: "Preparing multi-arm trials for meta-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing multi-arm trials for meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When meta-analyzing effect estimates obtained from multi-arm trials, we have to account for the dependence across estimates that arises when multiple treatment arms are compared to the same control group. The statistical methods for this have been worked out, as has an excellent software implementation in the `metafor` package. In short, we can pass a block-diagonal variance-covariance matrix to `rma.mv()` that appropriately accounts for this dependence.

The annoying part is making the block-diagonal variance-covariance matrix and keeping it synchronized with the estimates data frame. The `metaprep` package helps with this bookkeeping by providing an `estimates_vcov` object that keeps estimates and their variance-covariance matrix in sync through dplyr operations.

```{r setup, message=FALSE}
set.seed(343)
library(dplyr)
library(randomizr)
library(estimatr)
library(metaprep)
```

# Three experiments

Let's make fake data for three experiments. All three have a continuous outcome. The first is small (N = 50) with two arms. The second is larger (N = 100) with three arms. The third is the biggest (N = 200) with four arms. We use `randomizr::complete_ra()` to assign the treatment arm `Z`, which performs complete random assignment across the arms.

```{r}
make_trial_data <- function(n, n_arms) {
  arms <- paste0("T", seq_len(n_arms))
  data.frame(
    Z = complete_ra(N = n, conditions = arms),
    Y = rnorm(n)
  )
}

dat_1 <- make_trial_data(50, 2)
dat_2 <- make_trial_data(100, 3)
dat_3 <- make_trial_data(200, 4)
```

Now we estimate treatment effects with `estimatr::lm_robust()`, which returns heteroskedasticity-robust standard errors:

```{r}
fit_1 <- lm_robust(Y ~ Z, data = dat_1)
fit_2 <- lm_robust(Y ~ Z, data = dat_2)
fit_3 <- lm_robust(Y ~ Z, data = dat_3)
```

# Basic workflow: estimates_vcov objects

Here comes the `metaprep` part. We prep each fit, bind them together, and create an `estimates_vcov` object:

```{r}
prepped_fit_1 <- prep_fit(fit_1, term = "ZT2")
prepped_fit_2 <- prep_fit(fit_2, term = c("ZT2", "ZT3"))
prepped_fit_3 <- prep_fit(fit_3, term = c("ZT2", "ZT3", "ZT4"))

fits_df <- bind_rows(
  `Study 1` = prepped_fit_1,
  `Study 2` = prepped_fit_2,
  `Study 3` = prepped_fit_3,
  .id = "study"
)

ev <- as_estimates_vcov(fits_df)
ev
```

The `estimates_vcov` object keeps the estimates and variance-covariance matrix synchronized. You can use dplyr verbs on it:

```{r}
ev |> filter(study == "Study 2")
ev |> mutate(abs_estimate = abs(estimate))
ev |> arrange(desc(estimate))
```

Now we're ready to do meta-analysis using `rma_mv_helper()`:

```{r eval=requireNamespace("metafor", quietly = TRUE)}
library(metafor)
library(broom)
meta_fit <- ev |> rma_mv_helper(yi = estimate, random = ~ 1 | id)
tidy(meta_fit)
```

The `rma_mv_helper()` function is a convenience wrapper that automatically passes the estimates data frame and vcov matrix to `metafor::rma.mv()`. The `random = ~ 1 | id` argument specifies a random effect for each estimate (identified by the `id` column that `estimates_vcov` objects automatically include).

# Alternative workflow: extract components separately

If you prefer to work with the components separately, you can extract them:

```{r}
estimates_df <- get_estimates_df(fits_df)
estimates_df

vcov_matrix <- get_vcov(fits_df)
vcov_matrix
```

These functions work on both the prepped fits data frame and on `estimates_vcov` objects:

```{r}
estimates_df <- get_estimates_df(ev)
vcov_matrix <- get_vcov(ev)
```

You can also extract model-level summaries with `get_glance_df()`:

```{r}
glance_df <- get_glance_df(fits_df)
glance_df
```

Then conduct meta-analysis the traditional way:

```{r eval=requireNamespace("metafor", quietly = TRUE)}
meta_fit_traditional <- rma.mv(
  yi = estimate, V = vcov_matrix, data = estimates_df, random = ~ 1 | id
)
tidy(meta_fit_traditional)
```

# Grouping and nesting workflows

The `estimates_vcov` object supports `nest_by()` for grouped meta-analyses:

```{r eval=requireNamespace("metafor", quietly = TRUE)}
ev_grouped <- ev |>
  mutate(arm_type = if_else(term %in% c("ZT2"), "Treatment 2", "Treatment 3or4"))

grouped_results <- ev_grouped |>
  nest_by(arm_type) |>
  mutate(
    meta_fit = list(rma_mv_helper(data, yi = estimate, random = ~ 1 | id)),
    meta_results = list(tidy(meta_fit))
  ) |>
  tidyr::unnest(meta_results)

grouped_results |> select(arm_type, estimate, std.error, p.value)
```

# What's going on under the hood

`prep_fit()` conducts three post-estimation operations on the regression fit: `tidy()`, `glance()`, and `vcov()`. It subsets the rows of the tidy object and the rows and columns of the vcov object to the terms relevant for the meta-analysis -- in this case excluding the intercept.

The resulting three objects are wrapped in a tibble where the cells are themselves data frames:

```{r}
prepped_fit_1
```

We then bind all the prepped fits together with an id variable:

```{r}
fits_df
```

The `get_estimates_df()` function extracts and unnests the estimates:

```{r}
get_estimates_df(fits_df)
```

And the `get_vcov()` function extracts each `vcov_obj` and combines them into a block-diagonal matrix:

```{r}
get_vcov(fits_df)
```

The `as_estimates_vcov()` function creates a special object that keeps these two components synchronized:

```{r}
ev <- as_estimates_vcov(fits_df)
class(ev)
```

This object has an `id` column that links each row of estimates to the corresponding row/column of the vcov matrix:

```{r}
ev$estimates$id
rownames(ev$vcov)
```

When you use dplyr verbs on an `estimates_vcov` object, both the estimates and vcov are updated together:

```{r}
ev_study2 <- ev |> filter(study == "Study 2")
nrow(get_estimates_df(ev_study2))
dim(get_vcov(ev_study2))
```

This synchronization is what makes the `estimates_vcov` object useful for meta-analysis workflows involving filtering, grouping, and subsetting.
