---
title: "Preparing multi-arm trials for meta-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing multi-arm trials for meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When meta-analyzing effect estimates obtained from multi-arm trials, we have to account for the dependence across estimates that arises when multiple treatment arms are compared to the same control group. The statistical methods for this have been worked out, as has an excellent software implementation in the `metafor` package. In short, we can pass a block-diagonal variance-covariance matrix to the `rma.mv()` function that will appropriately account for this dependence.

The annoying part is making the block-diagonal variance covariance matrix and keeping it synchronized with the estimates data frame. The `metaprep` package helps with this bookkeeping by providing an `estimates_vcov` object that keeps estimates and their variance-covariance matrix in sync through dplyr operations.

```{r setup, message=FALSE}
set.seed(343)
library(tidyverse) # because we always need the tidyverse
library(randomizr) # to conduct random assignments
library(estimatr) # to estimate treatment effects and vcov
library(fabricatr)
library(metaprep) # preparing the estimates for meta-analysis 
library(metafor) # actually doing the meta-analysis
library(clubSandwich) # correction for small number of clusters
```

# Three experiments

Let's make fake data for three experiments. All three have a binary outcome. The first one is small (N = 50) and has two arms. The second is a little bigger (N = 100) and has three arms. The third is the biggest (N = 200) and has four arms.

```{r}
dat_1 <- 
  fabricate(
    N = 50,
    latent = rnorm(N),
    Y_Z_T1 = rbinom(n = N, 1, prob = pnorm(latent)),
    Y_Z_T2 = rbinom(n = N, 1, prob = pnorm(latent + 0.2)),
    Z = complete_ra(N = N, num_arms = 2),
    Y = reveal_outcomes(Y ~ Z)
  )

dat_2 <- 
  fabricate(
    N = 100,
    latent = rnorm(N),
    Y_Z_T1 = rbinom(n = N, 1, prob = pnorm(latent)),
    Y_Z_T2 = rbinom(n = N, 1, prob = pnorm(latent + 0.2)),
    Y_Z_T3 = rbinom(n = N, 1, prob = pnorm(latent + 0.3)),
    Z = complete_ra(N = N, num_arms = 3),
    Y = reveal_outcomes(Y ~ Z)
  )
dat_3 <- 
  fabricate(
    N = 200,
    latent = rnorm(N),
    Y_Z_T1 = rbinom(n = N, 1, prob = pnorm(latent)),
    Y_Z_T2 = rbinom(n = N, 1, prob = pnorm(latent + 0.2)),
    Y_Z_T3 = rbinom(n = N, 1, prob = pnorm(latent + 0.3)),
    Y_Z_T4 = rbinom(n = N, 1, prob = pnorm(latent + 0.2)),
    Z = complete_ra(N = N, num_arms = 4),
    Y = reveal_outcomes(Y ~ Z)
  )
```

Now we estimate treatment effects

```{r}
fit_1 <- lm_robust(Y ~ Z, data = dat_1)
fit_2 <- lm_robust(Y ~ Z, data = dat_2)
fit_3 <- lm_robust(Y ~ Z, data = dat_3)
```

# Basic workflow: estimates_vcov objects

Here comes the `metaprep` part. We prep each fit, bind them together, and create an `estimates_vcov` object:

```{r}
prepped_fit_1 <- prep_fit(fit_1, term = c("ZT2"))
prepped_fit_2 <- prep_fit(fit_2, term = c("ZT2", "ZT3"))
prepped_fit_3 <- prep_fit(fit_3, term = c("ZT2", "ZT3", "ZT4"))

fits_df <-
  bind_rows(
    `Study 1` = prepped_fit_1,
    `Study 2` = prepped_fit_2,
    `Study 3` = prepped_fit_3,
    .id = "study"
  )

# Create an estimates_vcov object
ev <- as_estimates_vcov(fits_df)
ev
```

The `estimates_vcov` object keeps the estimates and variance-covariance matrix synchronized. You can use dplyr verbs on it:

```{r}
# Filter keeps vcov in sync
ev |> filter(study == "Study 2")

# Mutate adds columns without affecting vcov
ev |> mutate(abs_estimate = abs(estimate))

# Arrange reorders both estimates and vcov
ev |> arrange(desc(estimate))
```

Now we're ready to do meta-analysis using the convenient `rma_mv_helper()`:

```{r}
meta_fit <- 
  ev |> 
  rma_mv_helper(yi = estimate, random = ~ 1 | id) |>
  robust(cluster = study, clubSandwich = TRUE)

tidy(meta_fit)
```

The `rma_mv_helper()` function is a convenience wrapper that automatically passes the estimates data frame and vcov matrix to `metafor::rma.mv()`. The `random = ~ 1 | id` argument specifies a random effect for each estimate (identified by the `id` column that `estimates_vcov` objects automatically include).

# Alternative workflow: extract components separately

If you prefer to work with the components separately, you can extract them:

```{r}
# Extract estimates data frame
estimates_df <- get_estimates_df(fits_df)
estimates_df

# Extract block-diagonal vcov matrix
vcov_matrix <- get_vcov(fits_df)
vcov_matrix
```

These functions work on both the prepped fits data frame and on `estimates_vcov` objects:

```{r}
# Works on estimates_vcov objects too
estimates_df <- get_estimates_df(ev)
estimates_df
vcov_matrix <- get_vcov(ev)
vcov_matrix
```

You can also extract model-level summaries with `get_glance_df()`:

```{r}
glance_df <- get_glance_df(fits_df)
glance_df
```

Then conduct meta-analysis the traditional way:

```{r}
meta_fit_traditional <- 
  rma.mv(yi = estimate, V = vcov_matrix, data = estimates_df, random = ~ 1 | id) |> 
  robust(cluster = study, clubSandwich = TRUE)

tidy(meta_fit_traditional)
```

# Grouping and nesting workflows

The `estimates_vcov` object supports `nest_by()` for grouped meta-analyses:

```{r}
# Add a grouping variable for demonstration
ev_grouped <- ev |> 
  mutate(arm_type = if_else(term %in% c("ZT2"), "Treatment 2", "Treatment 3or4"))

# Nest by group and run separate meta-analyses
grouped_results <- 
  ev_grouped |>
  nest_by(arm_type) |>
  mutate(
    meta_fit = list(rma_mv_helper(data, yi = estimate, random = ~ 1 | id)),
    meta_results = list(tidy(meta_fit))
  ) |>
  unnest(meta_results)

grouped_results |> 
  select(arm_type, estimate, std.error, p.value)
```

# What's going on under the hood

If you're interested, here's what's happening in these functions.

`prep_fit()` conducts three post-estimation functions on the regression fit: `tidy()`, `glance()`, and `vcov()`. It subsets the rows of the tidy object and the rows and columns of the vcov object to the "terms" that are relevant for the meta-analysis. In this case we're excluding the intercept. If the regressions adjusted for covariates, we could also exclude those.

The resulting three objects are wrapped in a tibble where the cells are themselves data frames:

```{r}
prepped_fit_1
```

We then bind all the prepped fits together with an id variable:

```{r}
fits_df
```

The `get_estimates_df()` function extracts and unnests the estimates:

```{r}
get_estimates_df(fits_df)
```

And the `get_vcov()` function extracts each `vcov_obj` and combines them into a block-diagonal matrix:

```{r}
get_vcov(fits_df)
```

The `as_estimates_vcov()` function creates a special object that keeps these two components synchronized:

```{r}
ev <- as_estimates_vcov(fits_df)
class(ev)
```

This object has an `id` column that links each row of estimates to the corresponding row/column of the vcov matrix:

```{r}
# The id column
ev$estimates$id

# Same ids on the vcov matrix
rownames(ev$vcov)
colnames(ev$vcov)
```

When you use dplyr verbs on an `estimates_vcov` object, both the estimates and vcov are updated together, maintaining synchronization:

```{r}
# Filter to just Study 2
ev_study2 <- ev |> filter(study == "Study 2")

# Estimates are filtered
nrow(get_estimates_df(ev_study2))

# And vcov is subset to match
dim(get_vcov(ev_study2))
```

This synchronization is what makes the `estimates_vcov` object so useful for meta-analysis workflows involving filtering, grouping, and subsetting.
